### 2025-08-10 base-rates.qmd
20250810 - edits

As we will see, the probability is: $\frac{95\% \times .3\%}{1\%} = 28.5\%$.
So based on the above probabilities, if you have a positive test, the probability that you have HIV is 28.5%.
Most people tend to vastly overestimate the likelihood that the person has HIV in this example.
Why?
Because they do not pay enough attention to the base rate (in this example, the base rate of HIV is .3%).

In general, people tend to overestimate the likelihood of low base-rate events.
That is, if the base rate of an event or condition—such as schizophrenia—is low (e.g., ~0.5%), people overestimate the likelihood that a person has schizophrenia when given specific information about the person such as their symptoms and history.
Moreover, @Fox1998a asked National Basketball Association (NBA) fans to estimate the probability that each of 8 teams in the playoff would win the playoffs.
The median sum of probability judgments for the eight teams was 240% [@Fox1998a; @Kahneman2011] even though, in reality, the true probabilities must sum to 100%.
The finding suggests that people were evaluating each team individually, considering the reasons why that particular team could win, without properly accounting for the total probability.

@@

@@

In sum, the [marginal probability](#sec-baseRate), including the [prior probability](#sec-baseRate) or [base rate](#sec-baseRate), should be weighed heavily in predictions unless there are sufficient data to indicate otherwise, i.e., to update the posterior probability based on new evidence.
People tend to ignore base rates (i.e., [base rate neglect](#sec-fallaciesBaseRate)).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
Bayes' theorem specifies how prior beliefs (i.e., [base rate](#sec-baseRate) informations) should be integrated with the [predictive accuracy](#sec-predictiveValidity) of the evidence to make predictions.
It thus provides a powerful tool to [anchor](#sec-heuristicsAnchoringAdjustment) predictions to the [base rate](#sec-baseRate) unless sufficient evidence changes the posterior probability (by updating the evidence and [prior probability](#sec-baseRate)).
In general, you should [anchor](#sec-heuristicsAnchoringAdjustment) your predictions to the [base rate](#sec-baseRate) and [adjust](#sec-heuristicsAnchoringAdjustment) from there.
As noted by @Kahneman2011, if you have doubts about the quality of the evidence for a particular prediction question, keep your predictions close to the [base rate](#sec-baseRate), and modify them only modifying them based on the new information.

@@

The positive likelihood ratio (LR+) compares the [true positive rate](#sec-sensitivity) to the [false positive rate](#sec-falsePositiveRate).
Positive likelihood ratio values range from 1 to infinity.\index{positive likelihood ratio}
Higher values reflect greater accuracy, because it indicates the degree to which a [true positive](#sec-truePositive) is more likely than a [false positive](#sec-falsePositive).
The formula for calculating the positive likelihood ratio is in @eq-positiveLikelihoodRatio.

@@

The negative likelihood ratio (LR−) compares the [false negative rate](#sec-falseNegativeRate) to the [true negative rate](#sec-specificity).
Negative likelihood ratio values range from 0 to 1.
Smaller values reflect greater accuracy, because it indicates that a [false negative](#sec-falseNegative) is less likely than a [true negative](#sec-trueNegative).
The formula for calculating the negative likelihood ratio is in @eq-negativeLikelihoodRatio.

>>>>

As we will see, the probability is: $\frac{95\% \times .3\%}{1\%} = 28.5\%$.
So based on the above probabilities, if you have a positive test, the probability that you have HIV is 28.5%.
Most people tend to vastly overestimate the likelihood that the person has HIV in this example.
Why?
Because they do not pay enough attention to the base rate (in this example, the base rate of HIV is .3%).
Base rates are important to account for when making predictions and people tend to ignore them, a phenomenon called [base rate neglect](#sec-fallaciesBaseRate), which is described in @sec-fallaciesBaseRate.

In general, people tend to overestimate the likelihood of low base-rate events [@Kahneman2011].
That is, if the base rate of an event or condition—such as schizophrenia—is low (e.g., ~0.5%), people overestimate the likelihood that a person has schizophrenia when given specific information about the person such as their symptoms and history.
As an example of people overestimating the likelihood of low base-rate events, @Fox1998a asked National Basketball Association (NBA) fans to estimate the probability that each of 8 teams in the playoff would win the playoffs.
The median sum of probability judgments for the eight teams was 240% [@Fox1998a; @Kahneman2011] even though, in reality, the true probabilities must sum to 100%.
The base rate for a given team to win is 12.5% (i.e., 1/8), whereas people were giving each team (on average) a 30% chance (i.e., 240/8) to win.
The finding suggests that people were evaluating each team individually, considering the reasons why that particular team could win, without properly accounting for the total probability (and the base rate).
In addition, people tend to overweight unlikely events in their decisions [@Kahneman2011].

Another important phenomenon is that low base-rate events (i.e., unlikely events) are difficult to predict accurately.
Predictions of lower base rate phenomena tend to be lower than predictions of more common phenomena.
For instance, predictions of touchdowns (less common) tend to be less accurate than predictions of yardage (more common).

In this chapter, we describe [ways to account for base rates](#sec-accountForBaseRates) in judgments and predictions.

@@

To calculate the probability of the inverse conditional, we can leverage Bayesian statistics to calculate the probability of the inverse conditional given a conditional probability (i.e., likelihood) and the marginal probabilities (base rates) of both events.
Bayesian statistics is another branch of statistics and is different from frequentist statistics and null hypothesis significance testing.
Bayesian statistics is based on Bayes’ theorem, which allows updating our probability estimates based on prior information (e.g., base rate) and new data.

@@

In sum, the [marginal probability](#sec-baseRate), including the [prior probability](#sec-baseRate) or [base rate](#sec-baseRate), should be weighed heavily in predictions unless there are sufficient data to indicate otherwise, i.e., to update the posterior probability based on new evidence.
People tend to ignore base rates (i.e., [base rate neglect](#sec-fallaciesBaseRate)).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
Bayes' theorem specifies how prior beliefs (i.e., [base rate](#sec-baseRate) information) should be integrated with the [predictive accuracy](#sec-predictiveValidity) of the evidence to make predictions.
It thus provides a powerful tool to [anchor](#sec-heuristicsAnchoringAdjustment) predictions to the [base rate](#sec-baseRate) unless sufficient evidence changes the posterior probability (by updating the evidence and [prior probability](#sec-baseRate)).
In general, you should [anchor](#sec-heuristicsAnchoringAdjustment) your predictions to the [base rate](#sec-baseRate) and [adjust](#sec-heuristicsAnchoringAdjustment) from there.
As noted by @Kahneman2011, if you have doubts about the quality of the evidence for a particular prediction question, keep your predictions close to the [base rate](#sec-baseRate), and modify them only modifying them based on the new information.

@@

The positive likelihood ratio (LR+) is the probability that a person with the disease tested positive for the disease ([true positive rate](#sec-sensitivity)) divided by the probability that a person without the disease tested positive for the disease ([false positive rate](#sec-falsePositiveRate)).
That is, the positive likelihood ratio (LR+) compares (i.e., is a ratio of) the [true positive rate](#sec-sensitivity) to the [false positive rate](#sec-falsePositiveRate).
Positive likelihood ratio values range from 1 to infinity.\index{positive likelihood ratio}
Higher values reflect greater accuracy, because it indicates the degree to which a [true positive](#sec-truePositive) is more likely than a [false positive](#sec-falsePositive).
Testing positive on a test with a high LR+ increases the probability of disease.
The formula for calculating the positive likelihood ratio is in @eq-positiveLikelihoodRatio.

@@

The negative likelihood ratio (LR−) is the probability that a person with the disease tested negative for the disease ([false negative rate](#sec-falseNegativeRate)) divided by the probability that a person without the disease tested negative for the disease ([true negative rate](#sec-specificity)).
That is, the negative likelihood ratio (LR−) compares (i.e., is a ratio of) the [false negative rate](#sec-falseNegativeRate) to the [true negative rate](#sec-specificity).
Negative likelihood ratio values range from 0 to 1.
Smaller values reflect greater accuracy, because it indicates that a [false negative](#sec-falseNegative) is less likely than a [true negative](#sec-trueNegative).
Testing negative on a test with a low LR– decreases the probability of disease.
The formula for calculating the negative likelihood ratio is in @eq-negativeLikelihoodRatio.
